A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-29 paint-contour-tree.tex
various changes

\begin{figure}[h]\centering
    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
    \hspace{.5in}
    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
    \caption{On the left, a surface will a balanced contour tree, but whose merge and split trees have long tails.  
    On the right (in clockwise order from the top) the contour, split and merge trees.}
    \label{fig:sorting}
\end{figure}

@@

A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
two disconnected complices.

\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
disconnected simplicial complices.
\end{theorem}

@@

In this section, we describe a linear time procedure that partitions $\MM$ into special
\emph{extremum dominant} simplicial complices.

@@

We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.

\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
results in two disjoint simplicial complices, one consisting entirely of dry points.
\end{claim}

\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complices. Let $\NN$ be the complex containing
the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
%
%
%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
%contradicting the fact that $\phi$ is dry.
\end{proof}

@@

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complices. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
of the procedure is that the rain alternately flows downwards and upwards, 
since otherwise faces may be cut a super constant number of times (see \Fig{order}).
In other words, when
we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
opposite direction. While one can implement $\rain$ this way, it is conceptually easier
to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
is easily achieved by just negating the height values. (From a running time standpoint, it suffices
to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

@@

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complices output by recursive calls.
\end{compactenum}
\end{minipage}}

@@

\begin{lemma}\label{lem:new-verts}
A facet $F \in \MM$ is cut at most once during $\rain(x,\MM)$.
\end{lemma}
\begin{proof} All notation here follows that in the pseudocode of $\rain$.
First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
Second, all recursive calls are made on dry complices.

@@

\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
complexities of the wet complices and the new boundaries created. No facet is ever wet twice,
and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
bounds the complexities of the boundaries created.
\end{proof}

@@

To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
from the join and split trees. 
It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
is the contour tree with regular points inserted into the edges. 
Consider tree $T$ with vertex $v$ of in and outdegree at most $1$.
\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.

@@

@@

\ignore{
\medskip
As the algorithm proceeds, it is convenient to imagine
it maintaining a set of contours $\Phi = \{\phi_1, \phi_2, \ldots, \phi_k\}$, which are contained
in incomparable edges of $\cC(\MM)$. Suppose we cut $\MM$ by each of these contours, and took
the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
have already determined the contour trees of the simplicial complices
where $\phi^+_i$ is a boundary maxima.

@@

@@

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
by edge $e$ of color $c$. Either all edges in $\PP$ are colored $c$, or there
exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
%
%and take $(f(v)+\delta)$-contour $\phi$ intersecting
%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
\end{lemma}

@@

 Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initial 
 adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (since elements are only removed 
 from heaps when being processed).  Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
 there is some vertex $w$ fully touched by $\rep(c)$ and some other color.  
 The former case implies that if there are any unprocessed vertices in $\PP$ then they are all in $T(\rep(c))$, 
 implying that $h$ is not the highest vertex.  Otherwise, all the vertices in $\PP$ have been processed.  
 However, it cannot be the case that all vertices in all components of $\MM^+_h$ have already been processed, 
 since this would imply that $h$ was ripe and so $h$ so can apply the same argument to the other non-fully processed component.

 Now consider the latter case, where we have a non-monochromatic vertex $w$.
 In this case $w$ cannot have been processed (since after being processed it is touched only by one color), 
 and so it must be in $T(\rep(c))$ since it must be in some heap of a color in each upstar (and one upstar is entirely colored $\rep(c)$).
 As $w$ lies above $h$ in $\MM$, this implies $h$ is not on the top of this heap.
\end{proof}

@@

\begin{proof} More generally, we prove the following: all outputs on processed vertices
are edges of $\jc(\MM)$ and for any color $c$, $\h(c)$ is the lowest processed vertex
of that color. We prove this by induction on the processing order. The base case is is trivially
true, as initially the processed vertices and attachments of the color classes are the set of maxima.
For the induction step, consider the situation
when $v$ is being processed.

@@

If there is another component of $\MM^+_v$, the same argument implies 
the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
Now by the definition of $\jc(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ where the lowest vertices before merging colors).

@@

@@

\Ben{All merge tree references need to be switched to join.  Also, need to make sure that not assuming the bin trees are full was safe to drop.}

In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
%
% \begin{definition}
%  We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
%  strictly increase along any root to leaf path.
% \end{definition}
% 
% \begin{observation}
%  The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
%  to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
%  factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
%  manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
% \end{observation}
%
As such any tree $T$ considered in this section will be a rooted binary tree\footnote{Note 
that technically the trees considered should have a leaf vertex hanging below the root of this 
in order to represent the global minimum of the complex.  This vertex is 
(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.
We use $A(v)$ to denote the set of ancestors of $v$.
For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.

\begin{definition}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
 one from the left child and one from the right child subtree of $v$.
\end{definition}

Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
according to the priority queue for each upstar we put a given vertex in.

\begin{lemma}
Let $\MM$ be a simplicial complex with $m$ edges.  For every vertex in $\jc(\MM)$, 
let $H_v$ be defined by the leaf assignment determined by $\init(\MM)$.
The running time of our algorithm is $O(m\alpha(m) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
% where $T$ is the merge tree of the given input, 
%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

\begin{proof}
First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(m)$ time (note that as the input complex is connected, $N=O(m)$).
Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e. each edge is viewed at most twice).
Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.

Now consider the union-find operations performed by $\init$ and $\update$.  
Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
and so each union-find operation takes $O(m)$ time.
The $\update(K)$ procedure performs at most one find for each edge and no unions.  
$\init(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
operations is $O(m)$.

For the remaining operations, observe that every time $\update$ is called a vertex is pushed onto the stack and each
vertex can only be pushed onto the stack once (since the only way it leaves the stack is by being processed). 
Therefore the total running time due to $\update$ is linear (ignoring the find operations).

What remains is the time it takes process a vertex $v$ in $\init(\MM)$.  
In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
Therefore the only thing we must bound are the queue operations.
Let $v$ be a vertex in $\jc(\MM)$, and let $p_1$ and $p_2$ be its parents (the same argument holds if $v$ has only one parent).
At the time $v$ is handled, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
In particular, when $v$ is handled we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
these merged queues are the queues of the current colors of $p_1$ and $p_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
the only vertices of left in these queues are those in $H_{p_1}$ and $H_{p_2}$.

Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $p_1$ and $p_2$, and then the queues are 
are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{p_1}$, $H_{p_2}$, and $H_v$, respectively.  
As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
\end{proof}

%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
%We will refer to $H_v$ as the \emph{heap} at $v$.
%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
%
%\begin{lemma}
%\label{lem:algCost}
%NEEDS TO BE WRITTEN:
%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
%\end{lemma}
%
%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
%of this section proving that this running time is optimal.

@@

Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.

@@

\section{Lower Bound by Path Decomposition}
In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.

\subsection{Merge Trees}
In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.

%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
\end{lemma}
\begin{proof}
 We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
 this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
 whose size is that described in the lemma statement.

@@

 Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
 diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
 diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
 (since all these saddles have a direct line of sight to this apex).  In particular, 
 this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.

 Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
 there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
 the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
 corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
\end{proof}

>>>>

\begin{figure}[h]\centering
    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
    \hspace{.5in}
    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
    \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
    On the right (in clockwise order from the top) the contour, split and join trees.}
    \label{fig:sorting}
\end{figure}

@@

A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
two disconnected complexes.

\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
disconnected simplicial complexes.
\end{theorem}

@@

In this section, we describe a linear time procedure that partitions $\MM$ into special
\emph{extremum dominant} simplicial complexes.

@@

We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.

\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
results in two disjoint simplicial complexes, one consisting entirely of dry points.
\end{claim}

\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
%
%
%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
%contradicting the fact that $\phi$ is dry.
\end{proof}

@@

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complexes. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
of the procedure is that the rain alternately flows downwards and upwards, 
since otherwise faces may be cut a super constant number of times (see \Fig{order}).
In other words, when
we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
opposite direction. While one can implement $\rain$ this way, it is conceptually easier
to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
is easily achieved by just negating the height values. (From a running time standpoint, it suffices
to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

@@

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
\end{compactenum}
\end{minipage}}

@@

\begin{lemma}\label{lem:new-verts}
A facet $F \in \MM$ is cut at most once during $\rain(x,\MM)$.
\end{lemma}
\begin{proof} All notation here follows that in the pseudocode of $\rain$.
First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
Second, all recursive calls are made on dry complexes.

@@

\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
complexities of the wet complexes and the new boundaries created. No facet is ever wet twice,
and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
bounds the complexities of the boundaries created.
\end{proof}

@@

Using the above lemma we can now prove that computing the contour tree for a minimum dominant manifold 
amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
merging procedure from~\cite{csa-cctad-00} that constructs the contour tree from the join and split trees. 
It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
is the contour tree with regular points inserted into the edges. 
Consider tree $T$ with vertex $v$ of in and outdegree at most $1$.
\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.

@@

\Ben{This figure will not show up in black and white printout.}

\begin{figure}[h]\centering
    \includegraphics[width=.27\linewidth]{figs/mounds}%
    \hspace{.35in}
    \includegraphics[width=.27\linewidth]{figs/mound2}%
    \hspace{.4in}
    \includegraphics[width=.22\linewidth]{figs/mound3}
    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
    \label{fig:colors}
\end{figure}

@@

\ignore{
\medskip
As the algorithm proceeds, it is convenient to imagine
it maintaining a set of contours $\Phi = \{\phi_1, \phi_2, \ldots, \phi_k\}$, which are contained
in incomparable edges of $\cC(\MM)$. Suppose we cut $\MM$ by each of these contours, and took
the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
have already determined the contour trees of the simplicial complexes
where $\phi^+_i$ is a boundary maxima.

@@

\begin{observation}
\label{obs:twoQueues}
 Each unprocessed vertex is always in exactly one queue of the colors in each up-star.  Specifically, 
 for a given up-star of a vertex $v$, $\init(\MM)$ puts $v$ into the queue of exactly one color, say $c$.  As time goes on this queue 
 may merge with other queues, but while unprocessed $v$ is only ever (and always) in the queue of $\rep(c)$,  
 since $v$ is never added to a new queue and is not removed until it is processed.  
 In particular, finding the queues of a vertex in $\update(K)$ requires at most two union find operations (assuming each vertex records its two colors from $\init(\MM)$)
\end{observation}

@@

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
by edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
%
%and take $(f(v)+\delta)$-contour $\phi$ intersecting
%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
\end{lemma}

@@

 Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initial 
 adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (see \Obs{twoQueues}).  
 Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
 there is some vertex $w$ fully touched by $\rep(c)$ and some other color.  
 The former case implies that if there are any unprocessed vertices in $\PP$ then they are all in $T(\rep(c))$, 
 implying that $h$ is not the highest vertex and a new higher up unprocessed vertex will be put on the stack for the next iteration of the while loop.  
 Otherwise, all the vertices in $\PP$ have been processed.  
 However, it cannot be the case that all vertices in all components of $\MM^+_h$ have already been processed, 
 since this would imply that $h$ was ripe, and so one can apply the same argument to the other non-fully processed component.

 Now consider the latter case, where we have a non-monochromatic vertex $w$.
 In this case $w$ cannot have been processed (since after being processed it is touched only by one color), 
 and so it must be in $T(\rep(c))$ since it must be in some heap of a color in each up-star (and one up-star is entirely colored $\rep(c)$).
 As $w$ lies above $h$ in $\MM$, this implies $h$ is not on the top of this heap.
\end{proof}

@@

\begin{proof} More generally, we prove the following: all outputs on processed vertices
are edges of $\jc(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
of that color. We prove this by induction on the processing order. The base case is is trivially
true, as initially the processed vertices and attachments of the color classes are the set of maxima.
For the induction step, consider the situation
when $v$ is being processed.

@@

If there is another component of $\MM^+_v$, the same argument implies 
the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
Now by the definition of $\jc(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).

@@

\subsection{Running Time}
\label{sec:runTime}

We now bound the running time of the above algorithm.  In the subsequent sections we will then provide a 
matching lower bound for the running time.  Therefore, it will be useful to set up some terminology that can 
be used consistently in both places.  Specifically, the lower bound proof will be a purely combinatorial 
statement on colored rooted trees, and so the terminology is of this form.

Any tree $T$ considered in following will be a rooted binary tree\footnote{Note 
that technically the trees considered should have a leaf vertex hanging below the root of this 
in order to represent the global minimum of the complex.  This vertex is 
(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
from the root $r$ (i.e. conceptually $T$ will be a join tree with $r$ at the bottom).  
As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.
We use $A(v)$ to denote the set of ancestors of $v$.
For a set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.

\begin{definition}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
 one from the left child and one from the right child subtree of $v$ (naturally if $v$ has only one child it is assigned 
 only one color).
\end{definition}

Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
according to the priority queue for each up-star we put a given vertex in.

\begin{lemma}
\label{lem:runTimeUpper}
Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\jc(\MM)$, 
let $H_v$ be defined by the leaf assignment determined by $\init(\MM)$.
The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
\end{lemma}

\begin{proof}
First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(N)$ time.
Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e. each edge is viewed at most twice).
Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.

Now consider the union-find operations performed by $\build$ and $\update$.  
Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
and so each union-find operation takes $O(\alpha(t))$ time.
For $\update$, by \Obs{twoQueues}, each iteration of the while loop requires a constant number of finds (and no unions).
Specifically, if a vertex is found to be ripe (and hence processed next) then these can be charged to that vertex.
If a vertex is not ripe, then these can be charged to the vertex put on the stack.
As each vertex is put on the stack or processed at most once, $\update$ performs $O(t)$ finds overall.
Finally, $\build(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
operations is $O(t)$.

For the remaining operations, observe that for every iteration of the loop in $\update$, a vertex is pushed onto the stack and each
vertex can only be pushed onto the stack once (since the only way it leaves the stack is by being processed). 
Therefore the total running time due to $\update$ is linear (ignoring the find operations).

What remains is the time it takes to process a vertex $v$ in $\build(\MM)$.  
In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
Therefore the only thing left to bound are the queue operations.
Let $v$ be a vertex in $\jc(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
At the time $v$ is processed, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
In particular, when $v$ is processed we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
these merged queues are the queues of the current colors of $c_1$ and $c_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
the only vertices left in these queues are those in $H_{c_1}$ and $H_{c_2}$.

Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $c_1$ and $c_2$, and then the queues are 
are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{c_1}$, $H_{c_2}$, and $H_v$, respectively.  
As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
\end{proof}

@@

In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
(this assumption simplifies the presentation but is not necessary).

\Ben{For now we are keeping the above full bin tree assumption. At some point this should be carefully removed.}

@@

Our main theorem is the following (which by \Lem{runTimeUpper} upper bounds the running time of our algorithm). We use $|p|$ to denote the number of vertices in $p$.

@@

\section{Lower Bound by Path Decomposition}
In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
we first consider the easier case of join trees, after which the lower bound for more general contour trees will easily follow.

\subsection{Join Trees}
In this section we show that any algorithm for computing the join tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.

%We say two shrubs $S_1$ and $S_2$ (of join trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $T$ be a join tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
\end{lemma}
\begin{proof}
 We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
 this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
 whose size is that described in the lemma statement.

@@

 Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
 diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
 diamond will appear contiguously on a root to leaf path in the join tree of the mesh, where the leaf corresponds to the maxima of the diamond 
 (since all these saddles have a direct line of sight to this apex).  In particular, 
 this implies that, regardless of the heights assigned to the platforms, the join tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.

 Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
 there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
 the mesh overall.  Also, each one of these instantiations will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
 corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the join tree.
\end{proof}

\begin{corollary}
 \label{cor:joinLB}
 Let $A(S)$ be any comparison based algorithm to compute the join tree of any input surface $S$ (whose contour tree is a join tree).  
 Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
 join tree of $S$.
\end{corollary}
\begin{proof}
 Given an input surface $S$, by the above lemma, any path decomposition $P$ of its join tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
 with distinct join trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
 on each diamond in the mesh, yields a different join tree.
 Therefore, any comparison based algorithm which computes the join tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
 have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

\subsection{Contour Trees}
We first generalize previous terms to the case of contour trees.
In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
and the height function defined over the vertex set is such that every internal vertex in either a split or join 
(i.e. at most $2$ neighbors lie above or below it).
